Geometric models overestimate lake depth due to imperfect slope measurement

### 2020-06-09 manuscript/manuscript.Rmd
cross-link figures,tables, and equations

# Optional: Enable a 'Draft' watermark on the document
#watermark: true

# Customize footer, eg by referencing the vignette
# footer_contents: "Agriculture-lake nutrient relationships"

@@

One approach for predicting lake depth involves a geometric model that assumes lake basins correspond to an idealized shape such as a cone, bowl, or an elliptic sinusoid \citep{hollisterPredictingMaximumLake2011, neumannMaximumDepthAverage1959, yigzawNewGlobalStorage2018}. All such geometric models for lake depth prediction involve implicit assumptions about the terms of geometric formulae. In the simplest case, where lakes basins are treated as cones (Eq \ref{eq1}, Figure \ref{fig1}), two assumptions are required to make depth predictions for all lakes: 1) that nearshore land slope is a representative proxy for in-lake slope and 2) that the distance to the center of the lake is a representative proxy for the distance to the deepest point of the lake (Figure 1). This cone model imposes the following fixed (i.e. geometric) relationship between slope and horizontal distance:
\begin{align}
  depth_{geometric} = tan(slope) * distance \label{eq1}
\end{align}

@@

\begin{figure}[H]
  \begin{center}
    \includegraphics[width=0.45\textwidth,keepaspectratio]{../figures/slope_diagram_new}
  \end{center}
  \caption{Diagram showing the relations between true (black) and proxy (orange) metrics of lake geometry. Geometric depth calculated via Equation 1 requires a single distance and slope metric.}\label{fig1}
\end{figure}

@@

To answer these questions, we extracted the maximum depth (hereafter referred to as “observed maximum depth”), in-lake slope, shape (i.e., concave versus convex), and distance to the deepest point, of approximately 5,000 lakes in the Northeastern and Midwestern US from bathymetric map data. First, we examined whether measures of lake geometry (in-lake slope and distance to the deepest point of lakes) were related to geometry proxies (nearshore land slope and distance to lake centers). Next, we computed geometric depth estimates (Equation 1) and offsets to these estimates using the random forest algorithm (Equation 2). Covariates used in offset modeling included a variety of lake, watershed, and hydrologic subbasin measures that are available for all lakes (Table 1). We examined differences in overall prediction error corresponding to different inputs to geometric depth as well as differences in the relative distribution of prediction error in lakes with different characteristics.

Given that, by definition, the distance proxy (distance to the center of the lake) must always be greater or equal to the true distance value (distance to the deepest point of the lake), we expect that the use of this proxy will lead to overestimation of lake depth (Figure 1). Furthermore we expect to see greater overestimation error in reservoirs as compared to natural lakes because many reservoirs are known to be drowned river valleys where the deepest point is close to the edge at the end of the reservoir (i.e. next to the dam) rather than in the center of the reservoir \citep{lanza1985interactions}. Similar to distance proxies, we expect that the use of a nearshore land slope proxy will lead to prediction error according to differences in cross-section shape (Figure 2) such that the depth of U-shaped (i.e. concave) lakes will be overpredicted whereas the depth of V-shaped (i.e. convex) lakes will be underpredicted. Finally, we expect that offsets to geometric depth estimates will be most strongly related to lake area and hydrologic subbasin variables as these measures have been influential in prior studies \citep{oliver2016prediction}.

@@

\begin{table}
\centering
\setlength\tabcolsep{1.5pt} % default value: 6pt
\begin{tabular}{lllll}
  \hline
  Variable & Median & Q25 & Q75 & n\\
  \hline
  Max depth (m) & 8.2 (7) & 4.6 (3.7) & 14 (12) & 4850 (17700)\\
  \textbf{Elevation (m)} & 300 (340) & 180 (210) & 400 (460) & 4850 (17700)\\
  \textbf{Area (ha)} & 55 (33) & 21 (11) & 140 (100) & 4850 (17700)\\
  \textbf{Island area (ha)} & 0 (0) & 0 (0) & 0.18 (0.076) & 4850 (17700)\\
  \textbf{Perimeter (m)} & 4400 (3500) & 2500 (1800) & 8100 (7300) & 4850 (17700)\\
  \textbf{Shoreline development} & 1.7 (1.7) & 1.4 (1.4) & 2.1 (2.2) & 4850 (17700)\\
  \textbf{Watershed-lake ratio} & 7.8 (10) & 3.9 (4.4) & 17 (28) & 4850 (17700)\\
  Distance to deepest point (m) & 180 (-) & 110 (-) & 290 (-) & 4850 (-)\\
  Distance to lake center (m) & 240 (-) & 160 (-) & 380 (-) & 4850 (-)\\
  In-lake slope (m/m) & 0.046 (-) & 0.024 (-) & 0.079 (-) & 4850 (-)\\
  Nearshore land slope (m/m) & 0.077 (-) & 0.051 (-) & 0.11 (-) & 4850 (-)\\  
  \bottomrule
  \end{tabular}
  \caption{Summary of lake characteristics for the present study (and for lakes in the contiguous United States from <LAGOSUS-Depth citation>). Predictor variables for computing random forest offsets (Eq 2) are printed in bold face. Dashes (-) indicate an identical sample size among this study and that of the contiguous United States.} \label{table1}
\end{table}

Covariates used in random forest modeling (Table 1, Equation 2) for lake elevation, area, island area, perimeter, shoreline development, watershed to lake area ratio, and hydrologic subbasin (i.e. HUC4s), were obtained from the LAGOS-US LOCUS database (Smith et al. In Prep). Shoreline development is a measure of lake perimeter shape defined as:

@@

2.2. Proxy evaluation
We conducted a qualitative assessment of whether or not proxy measures of lake geometry are representative of their true values by visual inspection (i.e. plotting each proxy measure against its corresponding true value) and by computing coefficients of determination ($R^2$). We further tested proxy measures by examining their effect on lake depth prediction error. Our approach involved several steps. In the first step, we computed a geometric estimate of lake depth using only geometry information (depth_geometric, Equation 1). In the second step, we fit a random forest model to predict observed (i.e. true) depth as a function of geometric depth along with several covariates available for all lakes (Table 1). The purpose of this random forest “offset” modeling was to more rigorously test our expectations regarding prediction error among different formulations of depth_geometric and among different lake types. Each of these steps were executed iteratively for each combination of true and proxy values of slope and distance (Table 2).

## Model description
### Geometric model 
We used a geometric model of lakes where basins are treated as cones with a fixed relationship between slope and distance (Equation 1). Note that Equation 1 is a geometric formula and has no intercept or “coefficients” and it produces a perfect depth estimate given true values of slope and distance. To use this model to predict the depth of all lakes, there is a necessary assumption that proxy slope and distance measures, which are available for all lakes, are representative of true slope and distance (Figure 1).

### Random forest models
Nearly all prior studies predicting lake depth using geometric models include a statistical or machine learning model “layer” or “offset” to boost predictive accuracy. For our purposes, this offset modeling enabled us to test our expectations that prediction error would be different among different formulations of depth_geometric and among different lake types. It also facilitated direct comparison against prior models of lake depth including those that are non-geometric. We generated an “offset” to geometric depth \citep[sensu][]{hollisterPredictingMaximumLake2011} using the random forest algorithm and the \texttt{ranger} R package \citep{wrightRangerFastImplementation2017} to predict observed maximum depth as a function of covariates including geometric maximum depth (from Equation 1) along with the lake elevation, area, perimeter, and ratio/index measures listed in Table 1:

@@

### Model comparisons
We tested model sensitivity to slope and distance proxies by generating multiple “geometric maximum depth” estimates from 3 different model runs using each of the possible metric combinations for Equation 1 (true slope - proxy distance, proxy slope - true distance, proxy slope - proxy distance). Prior to entry into Equation 1, we standardized proxy distances to have the same numeric range as their true counterpart. The purpose of this standardization was to prevent lakes with extremely long proxy distances from having an outsized impact on model evaluation metrics.

@@

Although proxy distance to lake center was often much larger in magnitude compared to the true distance to the deepest point of lakes’, they were strongly related ($R^2$ = 0.8). In contrast, proxy nearshore land slope and true in-lake slope were more weakly related ($R^2$ = 0.17). For slope measures, most lakes had lower magnitude (i.e. shallower) nearshore land slope compared to true in-lake slope (Figure 3). Taken together, these results suggest that proxy distance to the center of lakes is representative of true distance to the deepest point of lakes while proxy nearshore land slope is not representative of true in-lake slope. In addition to overall differences between slope and distance measures, we found differences in these relationships among lake shape classes. For example, in-lake slope and distance to the deepest point of the lake metrics were consistently larger in magnitude for convex lakes as compared to concave lakes (Figure S5). However,  there were not similar differences among slope and distance metrics for natural lakes versus reservoirs (Figure S5).

@@

The use of proxy nearshore land slope had a larger effect on model fit and prediction error than the use of proxy distance to lake center (Table 2). More specifically, the true slope - proxy distance model had a better fit ($R^2$ = 0.73) and lower prediction error (RMSE = 4.23m) compared to the proxy slope - true distance model ($R^2$ = 0.26, RMSE = 6.87m). Furthermore, analysis of model residuals showed overestimation of lake depth for concave lakes when models included a proxy slope measure (Figure 4). We observed similar but smaller overestimation depending on if a lake was classified as a reservoir rather than a natural lake (Figure 4).

\begin{table}
  \centering
  % \setlength\tabcolsep{1.5pt} % default value: 6pt
\begin{tabular}{llll}
  \hline
  slope & distance & rmse & rsq\\
  \hline
  true & true & - & -\\
  true & proxy & 4.2 m & 0.73\\
  proxy & true & 6.9 m & 0.26\\
  proxy & proxy & 6.6 m & 0.31\\
  \bottomrule
\end{tabular}
\caption{Model fit and predictive accuracy metrics (RMSE = root mean square error, $R^2$ = coefficient of determination) for all combinations of true (in-lake slope, distance to the deepest point of the lake) and proxy (nearshore land slope, distance to lake center) metrics.} \label{table2}
\end{table}

@@

The most important covariates in offset models were those relating to spatial location, lake area, and perimeter (Figure S6). Conversely, watershed metrics and lake elevation had little contribution to random forest model fit (Figure S6). The spatial location (i.e. HUC4) covariate was notably less importance in the true slope model compared to the two proxy slope models. Model importance calculations indicated that omitting a geometric max depth measure results in a 130\%, 60\%, or 50\% increase in mean square error depending on the formulation of geometric max depth in Eq 1 (Figure S6).

>>>>

# Optional: Enable a 'Draft' watermark on the document
watermark: true

# Customize footer, eg by referencing the vignette
footer_contents: "Geometric lake depth models and bathymetry"

@@

One approach for predicting lake depth involves a geometric model that assumes lake basins correspond to an idealized shape such as a cone, bowl, or an elliptic sinusoid \citep{hollisterPredictingMaximumLake2011, neumannMaximumDepthAverage1959, yigzawNewGlobalStorage2018}. All such geometric models for lake depth prediction involve implicit assumptions about the terms of geometric formulae. In the simplest case, where lakes basins are treated as cones (Eq \ref{eq1}, Figure \ref{fig1}), two assumptions are required to make depth predictions for all lakes: 1) that nearshore land slope is a representative proxy for in-lake slope and 2) that the distance to the center of the lake is a representative proxy for the distance to the deepest point of the lake (Figure \ref{fig1}). This cone model imposes the following fixed (i.e. geometric) relationship between slope and horizontal distance:
\begin{align}
  depth_{geometric} = tan(slope) * distance \label{eq1}
\end{align}

@@

\begin{figure}[H]
  \begin{center}
    \includegraphics[width=0.45\textwidth,keepaspectratio]{../figures/slope_diagram_new}
  \end{center}
  \caption{Diagram showing the relations between true (black) and proxy (orange) metrics of lake geometry. Geometric depth calculated via Equation \ref{eq1} requires a single distance and slope metric.}\label{fig1}
\end{figure}

@@

To answer these questions, we extracted the maximum depth (hereafter referred to as “observed maximum depth”), in-lake slope, shape (i.e., concave versus convex), and distance to the deepest point, of approximately 5,000 lakes in the Northeastern and Midwestern US from bathymetric map data. First, we examined whether measures of lake geometry (in-lake slope and distance to the deepest point of lakes) were related to geometry proxies (nearshore land slope and distance to lake centers). Next, we computed geometric depth estimates (Equation \ref{eq1}) and offsets to these estimates using the random forest algorithm (Equation \ref{eq2}). Covariates used in offset modeling included a variety of lake, watershed, and hydrologic subbasin measures that are available for all lakes (Table \ref{table1}). We examined differences in overall prediction error corresponding to different inputs to geometric depth as well as differences in the relative distribution of prediction error in lakes with different characteristics.

Given that, by definition, the distance proxy (distance to the center of the lake) must always be greater or equal to the true distance value (distance to the deepest point of the lake), we expect that the use of this proxy will lead to overestimation of lake depth (Figure \ref{fig1}). Furthermore we expect to see greater overestimation error in reservoirs as compared to natural lakes because many reservoirs are known to be drowned river valleys where the deepest point is close to the edge at the end of the reservoir (i.e. next to the dam) rather than in the center of the reservoir \citep{lanza1985interactions}. Similar to distance proxies, we expect that the use of a nearshore land slope proxy will lead to prediction error according to differences in cross-section shape (Figure \ref{fig2}) such that the depth of U-shaped (i.e. concave) lakes will be overpredicted whereas the depth of V-shaped (i.e. convex) lakes will be underpredicted. Finally, we expect that offsets to geometric depth estimates will be most strongly related to lake area and hydrologic subbasin variables as these measures have been influential in prior studies \citep{oliver2016prediction}.

@@

\begin{table}
\centering
\setlength\tabcolsep{1.5pt} % default value: 6pt
\begin{tabular}{lllll}
  \hline
  Variable & Median & Q25 & Q75 & n\\
  \hline
  Max depth (m) & 8.2 (7) & 4.6 (3.7) & 14 (12) & 4850 (17700)\\
  \textbf{Elevation (m)} & 300 (340) & 180 (210) & 400 (460) & 4850 (17700)\\
  \textbf{Area (ha)} & 55 (33) & 21 (11) & 140 (100) & 4850 (17700)\\
  \textbf{Island area (ha)} & 0 (0) & 0 (0) & 0.18 (0.076) & 4850 (17700)\\
  \textbf{Perimeter (m)} & 4400 (3500) & 2500 (1800) & 8100 (7300) & 4850 (17700)\\
  \textbf{Shoreline development} & 1.7 (1.7) & 1.4 (1.4) & 2.1 (2.2) & 4850 (17700)\\
  \textbf{Watershed-lake ratio} & 7.8 (10) & 3.9 (4.4) & 17 (28) & 4850 (17700)\\
  Distance to deepest point (m) & 180 (-) & 110 (-) & 290 (-) & 4850 (-)\\
  Distance to lake center (m) & 240 (-) & 160 (-) & 380 (-) & 4850 (-)\\
  In-lake slope (m/m) & 0.046 (-) & 0.024 (-) & 0.079 (-) & 4850 (-)\\
  Nearshore land slope (m/m) & 0.077 (-) & 0.051 (-) & 0.11 (-) & 4850 (-)\\  
  \bottomrule
  \end{tabular}
  \caption{Summary of lake characteristics for the present study (and for lakes in the contiguous United States from <LAGOSUS-Depth citation>). Predictor variables for computing random forest offsets (Eq \ref{eq2}) are printed in bold face. Dashes (-) indicate an identical sample size among this study and that of the contiguous United States.} \label{table1}
\end{table}

Covariates used in random forest modeling (Table \ref{table1}, Equation \ref{eq2}) for lake elevation, area, island area, perimeter, shoreline development, watershed to lake area ratio, and hydrologic subbasin (i.e. HUC4s), were obtained from the LAGOS-US LOCUS database (Smith et al. In Prep). Shoreline development is a measure of lake perimeter shape defined as:

@@

### Proxy evaluation
We conducted a qualitative assessment of whether or not proxy measures of lake geometry are representative of their true values by visual inspection (i.e. plotting each proxy measure against its corresponding true value) and by computing coefficients of determination ($R^2$). We further tested proxy measures by examining their effect on lake depth prediction error. Our approach involved several steps. In the first step, we computed a geometric estimate of lake depth using only geometry information (depth_geometric, Equation \ref{eq1}). In the second step, we fit a random forest model to predict observed (i.e. true) depth as a function of geometric depth along with several covariates available for all lakes (Table \ref{table1}). The purpose of this random forest “offset” modeling was to more rigorously test our expectations regarding prediction error among different formulations of depth_geometric and among different lake types. Each of these steps were executed iteratively for each combination of true and proxy values of slope and distance (Table \ref{table2}).

## Model description
### Geometric model 
We used a geometric model of lakes where basins are treated as cones with a fixed relationship between slope and distance (Equation \ref{eq1}). Note that Equation \ref{eq1} is a geometric formula and has no intercept or “coefficients” and it produces a perfect depth estimate given true values of slope and distance. To use this model to predict the depth of all lakes, there is a necessary assumption that proxy slope and distance measures, which are available for all lakes, are representative of true slope and distance (Figure \ref{fig1}).

### Random forest models
Nearly all prior studies predicting lake depth using geometric models include a statistical or machine learning model “layer” or “offset” to boost predictive accuracy. For our purposes, this offset modeling enabled us to test our expectations that prediction error would be different among different formulations of depth_geometric and among different lake types. It also facilitated direct comparison against prior models of lake depth including those that are non-geometric. We generated an “offset” to geometric depth \citep[sensu][]{hollisterPredictingMaximumLake2011} using the random forest algorithm and the \texttt{ranger} R package \citep{wrightRangerFastImplementation2017} to predict observed maximum depth as a function of covariates including geometric maximum depth (from Equation \ref{eq1}) along with the lake elevation, area, perimeter, and ratio/index measures listed in Table 1:

@@

### Model comparisons
We tested model sensitivity to slope and distance proxies by generating multiple “geometric maximum depth” estimates from 3 different model runs using each of the possible metric combinations for Equation (true slope - proxy distance, proxy slope - true distance, proxy slope - proxy distance). Prior to entry into Equation \ref{eq1}, we standardized proxy distances to have the same numeric range as their true counterpart. The purpose of this standardization was to prevent lakes with extremely long proxy distances from having an outsized impact on model evaluation metrics.

@@

Although proxy distance to lake center was often much larger in magnitude compared to the true distance to the deepest point of lakes’, they were strongly related ($R^2$ = 0.8). In contrast, proxy nearshore land slope and true in-lake slope were more weakly related ($R^2$ = 0.17). For slope measures, most lakes had lower magnitude (i.e. shallower) nearshore land slope compared to true in-lake slope (Figure \ref{fig3}). Taken together, these results suggest that proxy distance to the center of lakes is representative of true distance to the deepest point of lakes while proxy nearshore land slope is not representative of true in-lake slope. In addition to overall differences between slope and distance measures, we found differences in these relationships among lake shape classes. For example, in-lake slope and distance to the deepest point of the lake metrics were consistently larger in magnitude for convex lakes as compared to concave lakes (Figure S5). However,  there were not similar differences among slope and distance metrics for natural lakes versus reservoirs (Figure S5).

@@

The use of proxy nearshore land slope had a larger effect on model fit and prediction error than the use of proxy distance to lake center (Table \ref{table2}). More specifically, the true slope - proxy distance model had a better fit ($R^2$ = 0.73) and lower prediction error (RMSE = 4.23m) compared to the proxy slope - true distance model ($R^2$ = 0.26, RMSE = 6.87m). Furthermore, analysis of model residuals showed overestimation of lake depth for concave lakes when models included a proxy slope measure (Figure \ref{fig4}). We observed similar but smaller overestimation depending on if a lake was classified as a reservoir rather than a natural lake (Figure \ref{fig4}).

\begin{table}
  \centering
  % \setlength\tabcolsep{1.5pt} % default value: 6pt
\begin{tabular}{llll}
  \hline
  slope & distance & RMSE & $R^2$\\
  \hline
  true & true & - & -\\
  true & proxy & 4.2 m & 0.73\\
  proxy & true & 6.9 m & 0.26\\
  proxy & proxy & 6.6 m & 0.31\\
  \bottomrule
\end{tabular}
\caption{Model fit and predictive accuracy metrics (RMSE = root mean square error, $R^2$ = coefficient of determination) for all combinations of true (in-lake slope, distance to the deepest point of the lake) and proxy (nearshore land slope, distance to lake center) metrics.} \label{table2}
\end{table}

@@

The most important covariates in offset models were those relating to spatial location, lake area, and perimeter (Figure S6). Conversely, watershed metrics and lake elevation had little contribution to random forest model fit (Figure S6). The spatial location (i.e. HUC4) covariate was notably less importance in the true slope model compared to the two proxy slope models. Model importance calculations indicated that omitting a geometric max depth measure results in a 130\%, 60\%, or 50\% increase in mean square error depending on the formulation of geometric max depth in Eq \ref{eq1} (Figure S6).
